# Literate Programming with R Markdown

## Preliminaries

### What is literate programming?

[Literate programming](https://en.wikipedia.org/wiki/Literate_programming) is a
way of writing code and natural language prose.  It mixes the source code of
a computer program with an explanation of what the program is doing and the
results of running the program.

The principal advantages of this coding style are:
- it allows documentation/explanation of code to intermingle with the code itself,
  in a more fluid and accessible (to non-programmers) way than source code
  comments.
- it helps prevent the description of a code process and the code itself from
  diverging
- it allows a single document to contain the entirety of an analysis that is
  based partially in code and partially in prose

"Source code" sounds rather grand, but in linguistics this could be just the
extraction of counts from a corpus, or the generation of a graph.  You won't
have to worry about keeping track of your code separately from the documents
(slide decks, handouts, articles, ...) that reference it.

### Getting started

You should download the (free, open source) [RStudio](https://www.rstudio.com/)
IDE.  This provides a simple environment to work with R and R Markdown.

We'll be using the [knitr](http://yihui.name/knitr/) package to convert
Markdown documents to HTML.  (Not that you need to know that; most of the
details are silently handled by RStudio.)

Once you have installed and opened RStudio, go to the "File > New File > R
Markdown" menu item to get started (or open this file in RStudio's editor).

### What is Markdown?

A simple language that translates to HTML.  RStudio explains it better than I
could...open the program and browse the "Markdown Quick Reference" under the
"Help" menu.

Why Markdown and not LaTeX?  Markdown/HTML is a more
versatile format than PDF (more easily viewed on small devices, more easily
published on the web, less complicated to compile).  HTML also allows more
interactive features, some of which we will see in this document.

RStudio does supports R/LaTeX literate programming, via either Sweave or knitr
-- the latter is superior, but must be enabled in RStudio's preferences.

### Viewing your document

When you are ready to see the output, click the "Knit HTML" button in
the toolbar above RStudio's R markdown editor.  This will transform your
document to HTML and pop up a window to show you the output.

## Using R Markdown

### Code and results

The simplest way to use R Markdown is to embed *code chunks* in your
document.  These are delimited by three backtick characters (located
above the tab key on US keyboards).  By default, a code chunk prints its
source and output in the resulting document:

```{r}
12 * 12
```

The result is exactly what you would see in the R console when running
these commands, prefixed by `##` to turn it into an R comment.  (This
allows you to copy and paste the code and results into an R session
without the results confusing R.)

We might want to include some code without running it (for example to
discuss some incorrect algorithm that we will not use).  In that case,
we add a *chunk option* to the code chunk, telling knitr not to evaluate
that chunk.  In this case, we want `eval=FALSE`.  Thus we might say:
dividing by zero is a bad idea:

```{r eval=FALSE}
12 / 0
```

Notice we don't see the result of this code (which would be `Inf`) in
the output.  We can also ask to see only the result, and not the code:
the answer to life, the universe, and everything is:

```{r echo=FALSE}
6 * 7
```

### Running code in an R session

RStudio compiles your document in a separate R process.  This means that
you cannot see the variables that your code creates.  If you want to run
the code in your current R session, use the "Chunks" menu above the R
Markdown editor on the right to run a single chunk, or all chinks in the
document.

### Graphics

Graphical inclusion in your document should, in the simple case, just work:

```{r}
df <- data.frame(x = rnorm(10), y = rnorm(10))
library(ggplot2)
ggplot(aes(x = x, y = y), data = df) + geom_point()
```

If you want to control the width and height of the plot, you can use the
`fig.width` and `fig.height` options.  Let's make the same plot smaller
(notice we don't have to re-import the `ggplot2` library or define the
`df` variable...the code chunks are run sequentially, so those things
are already taken care of by the previous chunk):

```{r fig.width=4, fig.height=4}
ggplot(aes(x = x, y = y), data = df) + geom_point()
```

### Named code chunks

You can give code chunks a name:

```{r foo}
x <- 2
```

(This seems like useless trivia, but will become important in the
discussion of caching below)

### Caching

RStudio by default re-runs each code chunk every time it compiles your
document.  This can quickly make compilation a long process.  Luckily,
there is a caching option, which will save the result of a chunk for
future compilation, eliminating the need to re-run the code.  By way of
example, here is a code chunk which prints a random number.  It should
change on each recompilation of the document:

```{r}
rnorm(1)
```

On the other hand, this cached chunk will calculate a random number the
first time it is run, then continue to print the same random number on
each compilation.

```{r random-cache, cache=TRUE}
rnorm(1)
```

Knitr uses a very simple algorithm to determine if a code chunk needs to
be re-run: it looks at the source code of that chunk.  So it is not
smart enough to know that random number generation will give a different
result each time.  On the other hand, if we add a space at the beginning
of a line in a cached chunk, it will be re-run even though the meaning
of the code has not changed at all.

**Caching can be powerful, but also dangerous!** It's easy to be tricked
into thinking that we are getting new randomness each run, when in
fact we are not because of the cache

#### Caching all code chunks

Because the cache is often a useful time-saver, we might want to cache
all code chunks in a document.  We can do this by putting a code chunk
like this one at the beginning of the document:

```{r eval=FALSE}
opts_chunk$set(cache = TRUE)
```

We can also set other default options this way (perhaps we want to make
`fig.width` and `fig.height` 5 inches, rather than the default 7).
These options can be overridden at the individual chunk level.

#### Cache dependencies

As our document gets more complicated, we might want to have
dependencies on multiple code chunks.  A common motivation for this is
to separate data input/cleaning from plotting and modeling.  Here's a
sort of contrived example:

```{r read-data, cache=TRUE}
## Instead of random numbers, imagine that this code reads data
## from an external file, a sometimes-time-consuming process.
data <- data.frame(x = rnorm(10))
data$y <- x + rnorm(10, 0, 0.1)
```

```{r make-plot, cache=TRUE, dependson="read-data"}
## Similarly, imagine that this plot and the following model
## take a long time to produce
ggplot(aes(x = x, y = y), data = data) + geom_point()
```

```{r make-model, cache=TRUE, dependson="read-data"}
lm(y ~ x, data = data)
```

If we change one of the plot or model code chunks, only that chunk will
be re-run.  On the other hand, if we change the chunk which reads the
data, the data will be re-read and the plot and model re-generated based
on the new data.  This is because we have told knitr that the plot and
model depend on the data reading.  (Read the source code of this
document to see how this is done.)  This uses the code chunk names
discussed above.  We can make a code chunk depend on multiple ancestor
chunks, using the `c` function in R: `dependson=c("chunk-one",
"chunk-2")`.

## More advanced features

### Math

R Markdown can display mathematical formulas using a LaTeX-like syntax.
Here's the quadratic formula:

$$x = \frac{-b \pm \sqrt{b^2-4ac}}{2a}$$

For shorter formulas, using single dollar signs gives a more compact
rendering suitable for embedding in a paragraph.  Here's a compact
version of the quadratic formula: $x = \frac{-b \pm
\sqrt{b^2-4ac}}{2a}$.  Here's some more text to make this more like a
paragraph.  Etc., etc., you get the picture.

The underlying technology is called [Mathjax](http://www.mathjax.org/);
you can read documentation and examples at its website.

### Tables

Knitr includes a very simple HTML table generator.  Here it is in
action, using a very famous data set about irises (included with R and
thus used in many examples):

```{r results='asis'}
data(iris)
kable(head(iris))
```

Note (in this document's source) that we've used the chunk option
`results='asis'` to tell knitr to insert the result as raw HTML into the
document, rather then treating it as simple text.

This facility can form the basis of sophisticated functionality with
added Javascript code.  Try `vignette("datatables")` (with the `knitr`
library loaded in your R session) to see an example of this data set in
a sortable, searchable table.

There are also packages that allow the printing of models, data frame
summaries, etc. as HTML.  These include the packages
[`xtable`](http://cran.r-project.org/web/packages/xtable/),
[`stargazer`](http://cran.r-project.org/web/packages/stargazer/), and
[`texreg`](http://cran.r-project.org/web/packages/texreg/vignettes/v55i08.pdf).
Each one is sadly a bit different; you'll have to pick the one that most
closely matches your needs/preferences.  Generally speaking, `xtable` is
a more "general-purpose" converter of two-dimensional table-ish things
to HTML, whereas the other two specialize in regression models and try
to add useful things to them (for example AIC, BIC, $R^2$, ...).

[Here](http://conjugateprior.org/2013/03/r-to-latex-packages-coverage/)
is a website that compares the coverage of these packages on various
types of models.

Here are a couple of examples.  First we load the packages and create
some fake data:

```{r results='asis'}
suppressMessages(library(stargazer))
library(xtable)
options(xtable.type="html", xtable.html.table.attributes="")
data2 <- data.frame(x = rnorm(100), x2 = rnorm(100, 1, 0.1))
data2$y <- with(data2, x + 2 * x2 + rnorm(100, 0, 0.1))
```

The `stargazer` package is very fancy:

```{r results='asis'}
stargazer(lm(y ~ x + x2, data = data2), type = "html")
```

Whereas the `xtable` package gives us output very similar to what we see in the
R console:

```{r results='asis'}
xtable(lm(y ~ x + x2, data = data2))
```

Here's an example of a cross-tabulation:

```{r results='asis'}
data3 <- data.frame(a = sample(LETTERS[1:4], 100, replace = TRUE),
                    b = sample(LETTERS[1:4], 100, replace = TRUE))
xtable(table(data3$a, data3$b))
```

Each of these packages has many options to control the output, which you
can read about.  And each works with LaTeX as well as HTML/Markdown.

### Animation

We can make animated figures too, although they are a bit clunky by
default.  (For example, they will not show up properly in the RStudio
previewer, but rather have to be viewed in a real browser such as Chrome
or Firefox.)

```{r fig.show="animate"}
print(ggplot(aes(x = x, y = y), data = df) + geom_point(color = "red"))
print(ggplot(aes(x = x, y = y), data = df) + geom_point(color = "green"))
print(ggplot(aes(x = x, y = y), data = df) + geom_point(color = "blue"))
```

It's possible to get nicer animations, like the one on [this
page](http://taiyun.github.io/blog/2012/07/k-means/), but it takes some
fiddling.
[Here](https://github.com/yihui/knitr-examples/blob/master/108-animation-scianimator.Rmd)
are (not exactly complete) details on what is needed to make the
approach work properly.

## Concluding thoughts


